---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Stopgap estimates of PMT-relevant information

```{r}
#| label: libraries-constants
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "libraries_ts.r"))
source(here::here("r", "constants.r"))
source(here::here("r", "functions.r"))


fsemp <- readRDS(here::here("sites", "baseline", "pmtbase", "firmsize_byemp.rds"))
fss <- readRDS(here::here("sites", "baseline", "pmtbase", "firmsize_synthetic.rds"))

```

Estimates of private sector wages by county, industry, and more-detailed firmsize breakdown, using 2019 SUSB data, "synthetically" decomposed into finer firm-size groups using relatively old Business Employment Dynamics data https://www.bls.gov/news.release/cewfs.nr0.htm.

## Quarterly breadowns by ***employment*** size

```{r}
#| label: data-checks
#| include: false

fsemp |> 
  filter(naics == "10") |>
  filter(variable=="avgwage")


```

```{r}
#| label: empsize-prep
#| include: false

tabbase <- fsemp |> 
  # filter(naics != "10") |>
  filter(variable!="avgwage") |> 
  select(-c(total, e2099, e100499)) |> 
  relocate(sum, .after=e500p)


```


```{r}
#| label: empsize-tab-function
#| include: false


empsize_tab <- function(variable){
  
  tabdata <- tabbase |> 
    filter(variable==!!variable, naics=="10") |> 
    adorn_totals()|> 
    mutate(across(-c(fips, area, naics, naicsdesc, variable, sum),
                ~ .x / sum))
  
  vlab <- case_when(variable=="wages" ~ "wages ($ millions)",
                    variable=="estabs" ~ "establishments",
                    variable=="emp" ~ "employment",
                    TRUE ~ "ERROR")
  
  vscale <- case_when(variable=="wages" ~ 1e-3,
                    TRUE ~ 1)
  
  title <- paste0("% share of private sector ", vlab, " by area and employment size")

  tab <- tabdata |> 
    gt() |> 
    tab_header(
      title = html(title),
      subtitle = html("Based on SUSB and other sources")
    ) |>
    cols_label(
      elt20="< 20 employees",
      e2049="20-49 employees (estimated from 20-99)",
      e5099="50-99 employees (estimated from 20-99)",
      e100249="100-249 employees (estimated from 100-249)",
      e250499="250-499 employees (estimated from 100-249)",
      e500p="500+ employees",
      sum="SUSB sum (will be lower than QCEW)"
    ) |> 
    fmt_percent(columns=-c(fips, area, sum),
               decimals=1) |> 
    fmt_number(columns=sum,
             scale=vscale,
             decimals=0)
  
  tab
}

```


### Wages by employment size

```{r}
#| label: empsize-wages
#| include: true


tab <- empsize_tab("wages")
tab

```


```{r}
#| label: empsize-establishments
#| include: true

tab <- empsize_tab("estabs")
tab

```



```{r}
#| label: empsize-employment
#| include: true

tab <- empsize_tab("emp")
tab

```




## EXPERIMENTAL: Quarterly payroll breadowns by guessed-at wage sizes

```{r}
#| label: data-prep
#| include: false

df1 <- fss |> 
  filter(naics != "10", !egroup %in% c("total", "sum"), estabs > 0) |> 
  mutate(paygroup=cut(avgfirmqpayroll, c(0, 100000, 312500, 375000, 437500, Inf)))

cuts <- c(0, 100000, 312500, 375000, 437500, 500000, Inf)
df1 <- fss |> 
  filter(naics != "10", !egroup %in% c("total", "sum"), estabs > 0) |> 
  mutate(paygroup=cut(avgfirmqpayroll, cuts)) |> 
  summarise(across(c(estabs, emp, wages), sum),
            .by=c(fips, area, naics, naicsdesc, paygroup))
skim(df1)

areapaygroup <- df1 |> 
  summarise(across(c(estabs, emp, wages), sum),
            .by=c(fips, area, paygroup))


df1 |> filter(is.na(paygroup))

count(df1, paygroup)

```

### Breakdown by area and payroll size (ESTIMATED)

#### Wages by area

```{r}
#| label: wagetable
#| include: true

tabbase <- areapaygroup |> 
  select(fips, area, paygroup, wages) |> 
  arrange(paygroup) |> 
  pivot_wider(names_from = paygroup, values_from = wages, values_fill = 0) |> 
  mutate(sum=rowSums(across(-c(fips, area)))) |> 
  adorn_totals()

tabdata <- tabbase |> 
  mutate(across(-c(fips, area, sum), ~ .x / sum))

tab <- tabdata |> 
  gt() |> 
  tab_header(
    title = html("% share of private sector wages in area by quarterly payroll"),
    subtitle = html("Based on SUSB and other sources")
  ) |>
  fmt_percent(columns=-c(fips, area, sum),
             # scale=1e-9,
             decimals=1)

tab
  
```

### Establishments by area

```{r}
#| label: estabstable
#| include: true

tabbase <- areapaygroup |> 
  select(fips, area, paygroup, estabs) |> 
  arrange(paygroup) |> 
  pivot_wider(names_from = paygroup, values_from = estabs, values_fill = 0) |> 
  mutate(sum=rowSums(across(-c(fips, area)))) |> 
  adorn_totals()

tabdata <- tabbase |> 
  mutate(across(-c(fips, area, sum), ~ .x / sum))

tab <- tabdata |> 
  gt() |> 
  tab_header(
    title = html("% share of private sector establishments in area by quarterly payroll"),
    subtitle = html("Based on SUSB and other sources")
  ) |>
  fmt_percent(columns=-c(fips, area, sum),
             # scale=1e-9,
             decimals=1)

tab
  
```

### Wages by industry

```{r}
indpaygroup <- df1 |> 
  summarise(across(c(estabs, emp, wages), sum),
            .by=c(naics, naicsdesc, paygroup))

```

```{r}
#| label: wageindtable
#| include: true

tabbase <- indpaygroup |> 
  select(naics, naicsdesc, paygroup, wages) |> 
  arrange(paygroup) |> 
  pivot_wider(names_from = paygroup, values_from = wages, values_fill = 0) |> 
  mutate(sum=rowSums(across(-c(naics, naicsdesc)))) |> 
  adorn_totals() |> 
  arrange(desc(sum))

tabdata <- tabbase |> 
  mutate(across(-c(naics, naicsdesc, sum), ~ .x / sum)) |> 
  arrange(desc(sum))

tab <- tabdata |> 
  gt() |> 
  tab_header(
    title = html("% share of private sector wages in industries by quarterly payroll"),
    subtitle = html("Based on SUSB and other sources")
  ) |>
  fmt_percent(columns=-c(naics, naicsdesc, sum),
             # scale=1e-9,
             decimals=1)

tab
  
```

```{r}
#| label: estabsindtable
#| include: true

tabbase <- indpaygroup |> 
  select(naics, naicsdesc, paygroup, estabs) |> 
  arrange(paygroup) |> 
  pivot_wider(names_from = paygroup, values_from = estabs, values_fill = 0) |> 
  mutate(sum=rowSums(across(-c(naics, naicsdesc)))) |> 
  adorn_totals() |> 
  arrange(desc(sum))

tabdata <- tabbase |> 
  mutate(across(-c(naics, naicsdesc, sum), ~ .x / sum)) |> 
  arrange(desc(sum))

tab <- tabdata |> 
  gt() |> 
  tab_header(
    title = html("% share of private sector establishments in industries by quarterly payroll"),
    subtitle = html("Based on SUSB and other sources")
  ) |>
  fmt_percent(columns=-c(naics, naicsdesc, sum),
             # scale=1e-9,
             decimals=1)

tab
  
```
