---
output: html_document
editor_options: 
  chunk_output_type: console
---

# County allocation


## Questions for MTA

-   Do you care about intra-NYC distribution of taxes (e.g., Staten Island vs. Bronx)?

-   Reminder that it would be good to have any spreadsheets you can provide that underlie the MTA 2019 analysis [@metropolitantransportationauthorityMTACountyPayment2019]

## Preliminary allocation results

```{r}
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "constants.r"))
source(here::here("r", "functions.r"))

alloctotals_fn <- "090922 Subsidies for DB_djb.xlsx"

res <- readRDS(here::here("results", "allocation_results.rds"))

recipes <- read_excel(here::here("data", "mta", alloctotals_fn),
                      sheet="recipes",
                      range="A4:D23") |> 
  filter(!is.na(vname))

mtashares <- readRDS(here::here("data", "allocation", "allocators_mta.rds"))

# create a separate tibble of populations to make it easy for mta
pop <- res |> 
  select(geo_order, unifips, uniname, year, pop) |> 
  distinct() |> 
  arrange(geo_order, uniname, year) |> 
  select(-geo_order)
count(pop, unifips, uniname)

base_tabdata <- res |> 
  filter(!is.na(ftottax_summary)) |> 
  select(recipe, measure, year, geo_order, ftottax_summary, unifips, uniname,
         allocated, allocpc, ipcmta, pctofmta) |> 
  pivot_longer(c(allocated, allocpc, ipcmta, pctofmta)) |> 
  arrange(geo_order, uniname, ftottax_summary) |> 
  pivot_wider(names_from = ftottax_summary)


```

### 2021 accrual, recipe1,  \$ millions


```{r}
#| include: false

totals2021rec1 <- base_tabdata |> 
  filter(recipe=="recipe1", measure=="accrual", name=="allocated", year==2021) |> 
  arrange(geo_order, uniname)
  
tab <- totals2021rec1 |>   
  gt() |> 
  cols_hide(c(recipe, measure, year, geo_order, name)) |> 
  tab_header(
    title = "Preliminary allocation to counties, accruals, 2021",
    subtitle = "Amounts in $ millions, recipe1 (Surcharge: wages, not employment, by industry)"
  ) |> 
  cols_label(
    pmt = "Payroll mobility tax",
    mrt = "Mortgage recording tax (1&2)",
    urban="Urban tax",
    pbtall = "Petroleum business tax (incl.MMTOA)",
    optaxsum="Sum of operating taxes to left"
  ) |> 
  # tab_spanner(
  #   label = "Nonwage items",
  #   columns = c(supp_pct, propinc_pct)
  #   ) |> 
  # tab_spanner(
  #   label = "Wage items",
  #   columns = c(wk12_pct, wfed_pct, wsmall_pct)
  #   ) |> 
  # fmt_percent(
  #   columns=contains("pct"),
  #   decimals=1
  # ) |>
    fmt_number(
    columns=-c(unifips, uniname, year, recipe, measure, year, geo_order, name),
    scale=1e-6,
    decimals=1
  )

```

```{r}
#| include: true
tab
```



<!-- ### 2021 \$ per capita -->

<!-- ```{r} -->
<!-- #| include: false -->
<!-- tabdata <- res |>  -->
<!--   filter(rectype!="nyc") |>  -->
<!--   filter(year==2021, measure=="accrual") |>  -->
<!--   select(year, vname, unifips, uniname, allocpc) |>  -->
<!--   pivot_wider(names_from = vname, values_from = allocpc) |>  -->
<!--   mutate(mrt=mrt1 + mrt2) |>  -->
<!--   mutate(order=case_when(uniname=="MCTD" ~ 100, -->
<!--                          uniname=="New York City" ~ 99, -->
<!--                          uniname=="Suburban" ~ 98, -->
<!--                          TRUE ~ 1)) |>  -->
<!--   arrange(order, uniname) |>  -->
<!--   select(unifips, uniname, year, pmt, pbt, mrt, urban, mansion, -->
<!--          contains("mmtoa"), contains("mtaaid"), sutinternet) -->

<!-- tab <- tabdata |>  -->
<!--   gt() |>  -->
<!--   tab_header( -->
<!--     title = "Preliminary allocation to counties", -->
<!--     subtitle = "Per capita amounts" -->
<!--   ) |>  -->
<!--   cols_label( -->
<!--     pmt = "Payroll mobility tax", -->
<!--     pbt = "Petroleum business tax", -->
<!--     mrt = "Mortgage recording tax", -->
<!--     urban = "Urban tax" -->
<!--   ) |>  -->
<!--   # tab_spanner( -->
<!--   #   label = "Nonwage items", -->
<!--   #   columns = c(supp_pct, propinc_pct) -->
<!--   #   ) |>  -->
<!--   # tab_spanner( -->
<!--   #   label = "Wage items", -->
<!--   #   columns = c(wk12_pct, wfed_pct, wsmall_pct) -->
<!--   #   ) |>  -->
<!--   # fmt_percent( -->
<!--   #   columns=contains("pct"), -->
<!--   #   decimals=1 -->
<!--   # ) |> -->
<!--     fmt_number( -->
<!--     columns=-c(unifips, uniname, year), -->
<!--     scale=1, -->
<!--     decimals=1 -->
<!--   ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- #| include: true -->
<!-- tab -->
<!-- ``` -->


<!-- ### 2021 per capita indexed to overall MTA per-capita amount -->


<!-- ```{r} -->
<!-- #| include: false -->
<!-- tabdata <- res |>  -->
<!--   filter(rectype!="nyc") |>  -->
<!--   filter(year==2021, measure=="accrual") |>  -->
<!--   select(year, vname, unifips, uniname, ipcmta) |>  -->
<!--   pivot_wider(names_from = vname, values_from = ipcmta) |>  -->
<!--   mutate(mrt=mrt1 + mrt2) |>  -->
<!--   mutate(order=case_when(uniname=="MCTD" ~ 100, -->
<!--                          uniname=="New York City" ~ 99, -->
<!--                          uniname=="Suburban" ~ 98, -->
<!--                          TRUE ~ 1)) |>  -->
<!--   arrange(order, uniname) |>  -->
<!--   select(unifips, uniname, year, pmt, pbt, mrt, urban, mansion, -->
<!--          contains("mmtoa"), contains("mtaaid"), sutinternet) -->

<!-- tab <- tabdata |>  -->
<!--   gt() |>  -->
<!--   tab_header( -->
<!--     title = "Preliminary allocation to counties", -->
<!--     subtitle = "Indexed per capita amounts" -->
<!--   ) |>  -->
<!--   cols_label( -->
<!--     pmt = "Payroll mobility tax", -->
<!--     pbt = "Petroleum business tax", -->
<!--     mrt = "Mortgage recording tax", -->
<!--     urban = "Urban tax" -->
<!--   ) |>  -->
<!--   # tab_spanner( -->
<!--   #   label = "Nonwage items", -->
<!--   #   columns = c(supp_pct, propinc_pct) -->
<!--   #   ) |>  -->
<!--   # tab_spanner( -->
<!--   #   label = "Wage items", -->
<!--   #   columns = c(wk12_pct, wfed_pct, wsmall_pct) -->
<!--   #   ) |>  -->
<!--   # fmt_percent( -->
<!--   #   columns=contains("pct"), -->
<!--   #   decimals=1 -->
<!--   # ) |> -->
<!--     fmt_number( -->
<!--     columns=-c(unifips, uniname, year), -->
<!--     scale=1, -->
<!--     decimals=3 -->
<!--   ) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- #| include: true -->
<!-- tab -->
<!-- ``` -->



<!-- ### 2021 amounts as % of MTA total -->

<!-- ```{r} -->
<!-- #| include: false -->
<!-- tabdata <- res |>  -->
<!--   filter(rectype!="nyc") |>  -->
<!--   filter(year==2021, measure=="accrual") |>  -->
<!--   select(year, vname, unifips, uniname, allocated) |>  -->
<!--   pivot_wider(names_from = vname, values_from = allocated) |>  -->
<!--   mutate(mrt=mrt1 + mrt2) |>  -->
<!--   select(unifips, uniname, year, pmt, pbt, mrt, urban, mansion, -->
<!--          contains("mmtoa"), contains("mtaaid"), sutinternet) |>  -->
<!--   mutate(order=case_when(uniname=="MCTD" ~ 100, -->
<!--                          uniname=="New York City" ~ 99, -->
<!--                          uniname=="Suburban" ~ 98, -->
<!--                          TRUE ~ 1)) |>  -->
<!--   arrange(order, uniname) |>  -->
<!--   mutate(sum = rowSums(across(-c(unifips, uniname, year))), -->
<!--          across(-c(unifips, uniname, year), ~ .x / .x[uniname=="MCTD"])) |>  -->
<!--   select(-order) -->

<!-- tab <- tabdata |>  -->
<!--   gt() |>  -->
<!--   tab_header( -->
<!--     title = "Preliminary allocation to counties", -->
<!--     subtitle = "Amounts in $ millions" -->
<!--   ) |>  -->
<!--   cols_label( -->
<!--     pmt = "Payroll mobility tax", -->
<!--     pbt = "Petroleum business tax", -->
<!--     mrt = "Mortgage recording tax", -->
<!--     urban = "Urban tax" -->
<!--   ) |>  -->
<!--   # tab_spanner( -->
<!--   #   label = "Nonwage items", -->
<!--   #   columns = c(supp_pct, propinc_pct) -->
<!--   #   ) |>  -->
<!--   # tab_spanner( -->
<!--   #   label = "Wage items", -->
<!--   #   columns = c(wk12_pct, wfed_pct, wsmall_pct) -->
<!--   #   ) |>  -->
<!--   # fmt_percent( -->
<!--   #   columns=contains("pct"), -->
<!--   #   decimals=1 -->
<!--   # ) |> -->
<!--     fmt_percent( -->
<!--     columns=-c(unifips, uniname, year), -->
<!--     decimals=1 -->
<!--   ) -->

<!-- ``` -->

<!-- ```{r} -->
<!-- #| include: true -->
<!-- tab -->
<!-- ``` -->


## Save results to excel

You won't see anything below. There is some hidden code that writes key information to an Excel file.
Given the time crunch I haven't bothered to format anything in the workbook

```{r}

outpath <- here::here("results", "county_allocation.xlsx")

f <- function(sheetname){
  setColWidths(wb,
               sheet=sheetname,
               cols=1:50,
               widths = 15, # 8.43,
               hidden = rep(FALSE, length(cols)),
               ignoreMergedCells = FALSE
  )
}

sheetlist <- list(recipes = recipes, population=pop, allocators=mtashares, 
                  totals2021rec1=totals2021rec1,
                  base_tabledata=base_tabdata,
                  allresults=res)

wb <- buildWorkbook(sheetlist, asTable = TRUE)

purrr::walk(names(sheetlist), f)

freezePane(wb, sheet="recipes", firstActiveRow = 2, firstActiveCol = 2)
freezePane(wb, sheet="population", firstActiveRow = 2, firstActiveCol = 4)
freezePane(wb, sheet="allocators", firstActiveRow = 2, firstActiveCol = 7)
freezePane(wb, sheet="totals2021rec1", firstActiveRow = 2, firstActiveCol = 8)
freezePane(wb, sheet="base_tabledata", firstActiveRow = 2, firstActiveCol = 8)
freezePane(wb, sheet="allresults", firstActiveRow = 2, firstActiveCol = 9)

saveWorkbook(wb, file = outpath, overwrite = TRUE)

# view workbook -----------------------------------------------------------

# openXL(wb)  ## opens a temp version


# 
# # define styles -----------------------------------------------------------
# num0 <- createStyle(numFmt = "#,##0")
# num1 <- createStyle(numFmt = "#,##0.0")
# num2 <- createStyle(numFmt = "#,##0.00")
# num3 <- createStyle(numFmt = "#,##0.000")
# 
# curr0 <- createStyle(numFmt = "$#,##0")
# curr2 <- createStyle(numFmt = "$#,##0.00")
# curr3 <- createStyle(numFmt = "$#,##0.000")
# 
# pct1 <- createStyle(numFmt = "0.0%")
# pct2 <- createStyle(numFmt = "0.00%")
# 
# # for totals rows
# totstyle <- createStyle(fontSize = 12, fgFill = "#f0f0f0", border = "TopBottom", borderColour = "#f0f0f0")
# headerstyle <- createStyle(fontSize = 12, fgFill = "#f0f0f0", 
#                            border = "TopBottom", borderColour = "#f0f0f0", 
#                            wrapText = TRUE, textDecoration = "bold")
# bold_style <- createStyle(fontSize = 12, textDecoration = "bold")
# wrap_style <- createStyle(wrapText = TRUE, halign = "left", valign = "center")
# vcenter_style <- createStyle(valign = "center")
# 
# 
# # get empty workbook shell ------------------------------------------------------
# shellfile <- "nonprofits_shell.xlsx"
# snames <- getSheetNames(here::here("report", "inputs", shellfile))
# 
# wb <- loadWorkbook(file = here::here("report", "inputs", shellfile))
# 
# # global workbook settings
# modifyBaseFont(wb, fontSize = 12, fontColour = "black", fontName = "Calibri")
# 
# 
# # functions ---------------------------------------------------------------
# 
# top_sheet <- function(wb, data, sname, first_data_row,
#                       comma_cols, dollar_cols, pct_cols,
#                       total_format=TRUE){
#   last_data_row <- nrow(data) + first_data_row - 1
#   totalrow <- last_data_row
#   last_col <- ncol(data)
#   
#   writeData(wb, sheet = sname, x=data,
#             startCol=1, startRow=first_data_row, rowNames = FALSE, colNames = FALSE)
#   
#   # apply styles
#   addStyle(wb, sheet = sname, style = num0, rows=first_data_row:last_data_row, cols = comma_cols, gridExpand = TRUE)
#   addStyle(wb, sheet = sname, style = pct1, rows=first_data_row:last_data_row, cols = pct_cols, gridExpand = TRUE)
#   addStyle(wb, sheet = sname, style = curr0, rows=first_data_row, cols = dollar_cols)
#   
#   if(total_format){
#     # addStyle(wb, sheet = sname, style = curr0, rows=last_data_row, cols = dollar_cols)
#     addStyle(wb, sheet = sname, style = curr0, rows=totalrow, cols = dollar_cols, stack=TRUE)
#     addStyle(wb, sheet = sname, style = totstyle, rows = totalrow, cols = 1:last_col, stack=TRUE)
#   }
#   
#   wb
# }
# 
# 
# npes_by_depts <- function(wb, data, sname, header_row){
#   
#   cnames <- c("Rank", "Nonprofit Organization", names(data)[-c(1:2)])
#   # create an empty data frame with these names
#   cnamesdf <- as_tibble(matrix(nrow=0, ncol=length(cnames), dimnames=list(NULL, cnames)))
#   
#   first_data_row <- header_row + 2
#   last_data_row <- nrow(data) + first_data_row - 1
#   totalrow <- last_data_row
#   
#   last_col <- ncol(data)
#   comma_cols <- 3:last_col
#   dollar_cols <- comma_cols
#   
#   # clear all data
#   # deleteData(wb, sheet = sname, rows = header_row:totalrow, cols = 1:last_col, gridExpand = TRUE)
#   # removeCellMerge(wb, sheet = sname, rows = 1:totalrow, cols = 1:last_col)
#   
#   # insert column names
#   writeData(wb, sheet = sname, x=cnamesdf,
#             startCol=1, startRow=header_row, rowNames = FALSE, colNames = TRUE)
#   # data
#   writeData(wb, sheet = sname, x=data,
#             startCol=1, startRow=first_data_row, rowNames = FALSE, colNames = FALSE)
#   
#   # apply styles
#   setColWidths(wb, sheet=sname, cols = comma_cols, widths = rep(15, length(comma_cols)))
#   mergeCells(wb, sheet = sname, rows = 3, cols = 1:last_col)
#   mergeCells(wb, sheet = sname, rows = 4, cols = 1:last_col)
#   
#   addStyle(wb, sheet = sname, style = headerstyle, rows = header_row, cols = 1:last_col, stack=TRUE)
#   
#   addStyle(wb, sheet = sname, style = num0, rows=first_data_row:last_data_row, cols = comma_cols, gridExpand = TRUE)
#   addStyle(wb, sheet = sname, style = curr0, rows=c(first_data_row, last_data_row), cols = dollar_cols, gridExpand = TRUE)
#   
#   addStyle(wb, sheet = sname, style = curr0, rows=totalrow, cols = dollar_cols, stack=TRUE)
#   addStyle(wb, sheet = sname, style = totstyle, rows = totalrow, cols = 1:last_col, stack=TRUE)
#   wb
# }

```



